---
title: "Prepare-data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prepare-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE}
library(PEMr)
library(PEMprepr)

#project_name <- "{{ aoi_name }}"
#project_name <- "datecreek_aoi"

fid <- read_fid()

```

# create templates 

```{r, eval = FALSE}

snap_aoi()

create_template_raster()


```


# generate base vector layers

```{r, eval = FALSE}
create_base_vectors()

```


# generate base DEM layer for modelling (5m)

```{r, eval = FALSE}

#1) option to use trim data 

get_cded_dem()

#2) option to use lidr 


```

# generate raster covariates for modelling (5m)


```{r}

create_covariates()

```


# convert bec layer into raster


Create a bec raster to match the template sizes 

```{r}
# 
# bec_sf <- sf::st_read(file.path(fid$shape_dir_1010[1], "bec.gpkg")) %>%
#   sf::st_cast(., "MULTIPOLYGON") 
# 
# bec_code <- bec_sf %>% st_drop_geometry()  %>% dplyr::select(MAP_LABEL) %>%
#     unique() 
# 
# bec_code <- bec_code %>% 
#     mutate(bgc_unique_code = seq(1, length(bec_code$MAP_LABEL),1))
# 
# bec_sf <- dplyr::left_join(bec_sf, bec_code)
# 
# 
# bec_vec <- terra::vect(bec_sf)
# 
# # generate a 25m raster
# 
# bec_ras25 <- terra::rasterize(bec_vec, r25, field = "MAP_LABEL")
# 
# terra::writeRaster(bec_ras25, file.path(fid$cov_dir_1020[2], "25m", "bec.tif"), overwrite = TRUE)
# 
# 
# # generate a 5m raster 
# 
# bec_ras5 <- terra::rasterize(bec_vec, r5, field = "MAP_LABEL")
# 
# terra::writeRaster(bec_ras5, file.path(fid$cov_dir_1020[2], "5m", "bec.tif"), overwrite = TRUE)

```








# generate inputs at landscape level (25m resolution)

# generate base landscape template raster 

```{r, eval = FALSE}

create_raster_template(res = 25)

```

# generate dem at 25m res

```{r}
get_cded_dem(res = 25)
```

# landscape polygon 

```{r}

create_covariates()

```




# Create the suplementary data. 
## Create fuzzy matrix 

```{r}

#create_fuzzy_matrix()

```

