---
title: "Run_nonforest_model_07"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run_nonforest_model_07}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PEMr)
```


# Generate Non-Forest model

This workflow produces a model for non-forest groupings. As the sample design was specifically designed for forested landscapes this workflow provides a very basic non-forest model. This could be substituted with other non-forest mapping products

We provide this workflow as a placeholder for non-forest mapping products.


## 1.1 prep all training points

Prepare training points for using the attribute "mapunit_nf" in the mapkey.csv. This differs to the forest only mapping. 

As previously we will generate a full list of covariates. 

```{r}
#| eval: false

out_dir = fs::path(PEMprepr::read_fid()$dir_3020_draft$path_rel, "30_nf")
covarkey = read.csv(fs::path(PEMprepr::read_fid()$dir_30_model$path_rel,"covar_key.csv"))

tpts <- prep_tps(
  allpts = sf::st_read(fs::path(PEMprepr::read_fid()$dir_20105030_attributed_field_data$path_rel, "allpoints_att.gpkg")),
  mapkey = read.csv(fs::path(PEMprepr::read_fid()$dir_30_model$path_rel,"mapunitkey_final.csv")),
  covarkey = read.csv(fs::path(PEMprepr::read_fid()$dir_30_model$path_rel,"covar_key.csv")),
  attribute = "mapunit_nf1",
  extra_pts = TRUE,
  bec = sf::st_read(fs::path(PEMprepr::read_fid()$dir_1010_vector$path_rel, "bec.gpkg")),
  min_no = 20)

core_names  <- covarkey |>
  dplyr::filter(type == "core") |>
  dplyr::select(value) |> dplyr::pull()


# get full list of covariates
mcols <- names(tpts)[!names(tpts) %in% c(core_names)]
saveRDS(mcols, fs::path(out_dir, "full_covariate_list.rds"))


# convert to csv
tpts <- cbind(tpts, as.data.frame(sf::st_coordinates(tpts))) |>
  sf::st_drop_geometry()

write.csv(tpts, fs::path(out_dir, "training_pts.csv"), row.names = FALSE)

```



## 1.2) Generate a fuzzy matrix. 

To assist in accuracy assessment we need to generate a fuzzy metrics. This provides a list of all mapunits  we need to create a table which assigns partial correct values for mapunit calls that are similar (on the edatopic position) to the correct calls. In this case, scores could be awarded for on a sliding scale from 1 (Correct) to 0 (nowhere close) with partial credit assigned to closely related mapunits. Note this requires a matrix which specifies the similarity between all combinations of possible calls. 

The final output is a list of all mapunits within the tpts$mapunit1 column and cross reference with all other units with an equivalent value between 0 and 1. The dataframe consists of three columns: target, Pred, f. 

Note for this model the types include non-forest so we have a simple fuzzy matrix. Might not need this as "nonforest and forest cross over = 0"

```{r}
#| eval: false
#| 
#| 
fmat <- generate_fuzzymatrix(tpts, 
                             "bgc_cat", 
                             out_dir = out_dir,
                             out_name = "fuzzy_matrix.csv",
                             write_output = FALSE,
                             overwrite = TRUE)

fmat <- read.csv(fs::path(out_dir, "fuzzy_matrix.csv" ))

```

Note a warning message will show if not all units are included in the fuzzy matrix. This will need to be manually checked to ensure all units are included or the model runs will throw errors.



## 1.3)  Recursive Feature selection / Correlated variable reduction

To improve the interpretability of the models, we can firstly remove the correlated covariates. While there are several methods available to complete this task we chose a simple correlation matrix with a cutoff values of 0.9.  This is based on the entire raster surface (not just the points). The cutoff parameters can be adjusted. 

In the forest/non-forest models we also use the covariates grouped as "dem",  "structure" and "satellite".

```{r}
#| eval: false
mcols <- readRDS(fs::path(out_dir, "full_covariate_list.rds"))

reduced_vars <- reduce_features(mcols,
                                covar_dir = fs::path(PEMprepr::read_fid()$dir_1020_covariates$path_rel, "5m"),
                                covarkey = covarkey,
                                covtype = c("dem", "structure", "satellite"),
                                cutoff = 0.90)

write.csv(reduced_vars, fs::path(out_dir, "reduced_covariate_list.csv"), row.names = FALSE)
```



## 1.4) prepare the training point models (fnf)

```{r}
#| eval: false

tpts <- read.csv(fs::path(out_dir, "training_pts.csv"))

model_bgc <- prep_model_tps(
  prepped_points = tpts,
  covars = reduced_vars,
  model_type = "nf",
  outname = "model_input_pts.rds",
  out_dir = out_dir
)

```


# 1.5) determine optimum parameters for the model

We use only the original data points (not the neighbours) and pure calls to tune the model hyperparameters. This includes estimating mtry and min_n to use for the models. Best accuracy is determined based on the roc_auc and accuracy metrics. We can also plot the each is exported to enable more assessment of best options. 
```{r}
#| eval: false

tune_bgc <- tune_model_params(
  prepped_points = model_bgc,
  covars = reduced_vars,
  min_no = 20,
  out_dir = out_dir,
  output_type = "best",
  accuracy_type = "roc_auc"
)

```

## 2) Run the base model

```{r}
#| eval: false
out_dir = fs::path(PEMprepr::read_fid()$dir_3020_draft$path_rel, "30_nf")

bgc_pts_subzone <- readRDS(fs::path(out_dir, "model_input_pts.rds"))

fuzz_matrix <- read.csv(file.path(out_dir, "fuzzy_matrix.csv" )) 

covars = read.csv(fs::path(out_dir, "reduced_covariate_list.csv")) |> dplyr::pull()


## Run base model
run_base_model(
  bgc_pts_subzone = bgc_pts_subzone,
  fuzz_matrix = fuzz_matrix,
  covars = covars,
  use_neighbours = FALSE,
  extra_pts = TRUE,
  report = FALSE,
  detailed_output = TRUE,
  out_dir = out_dir)


```



# 4) Run final model

In this example we use the base model as the final model. This could also be a balanced option. 

```{r}
#| eval: false

out_dir = fs::path(PEMprepr::read_fid()$dir_3020_draft$path_rel, "30_nf")
train_data <- readRDS(fs::path(out_dir, "model_input_pts.rds"))
covars <- read.csv(fs::path(out_dir, "reduced_covariate_list.csv")) |> dplyr::pull()


final_mod <- run_final_model(train_data,
                             covars,
                             model_bal = "base",
                             report = FALSE,
                             extra_pts = TRUE,
                             out_dir = out_dir)

```


# 5) Predict maps

```{r}
#| eval: false

model_dir = fs::path(PEMprepr::read_fid()$dir_3020_draft$path_rel, "30_nf")
covars = read.csv(fs::path(model_dir, "reduced_covariate_list.csv")) |>  dplyr::pull()
cov_dir = fs::path(PEMprepr::read_fid()$dir_1020_covariates$path_rel, "5m")
tile_dir = fs::path(PEMprepr::read_fid()$dir_30_model$path_rel,"tiles")
bec_shp <- sf::st_read(fs::path(PEMprepr::read_fid()$dir_1010_vector$path_rel,"bec.gpkg"), quiet = TRUE)

run_predict_map(
  model_type = "nf",
  model_dir = model_dir,
  model_name = "final_model_base.rds",
  covars = covars,
  cov_dir = cov_dir,
  tile_dir = tile_dir,
  bec_shp = bec_shp
)
```


Advanced modeling options (optional)

```{r}
#| eval: false

# 4.1 Determine optimum Theta value for the metric of choice
# run all theta options and consolidate output

out_dir = fs::path(PEMprepr::read_fid()$dir_3020_draft$path_rel, "30_nf")

# read in data
bgc_pts_subzone <- readRDS(fs::path(out_dir, "model_input_pts.rds"))

# read in fuzzy matrix
fuzz_matrix <- read.csv(file.path(out_dir, "fuzzy_matrix.csv" ))

run_theta_metrics(bgc_pts_subzone, out_dir, fuzz_matrix, overwrite = TRUE)


############################################################

# 4.2 Determine to optimum Balance options for models

############################################################

## Build grid of combinations of downsampling and smoting and run all
## combinations then compare outputs to assess optimum
## Note optimum will depend on which outcome desired.

out_dir = fs::path(PEMprepr::read_fid()$dir_3020_draft$path_rel, "30_nf")

fuzz_matrix <- read.csv(fs::path(out_dir , "fuzzy_matrix.csv" ))

covars = read.csv(fs::path(out_dir , "reduced_covariate_list.csv")) |> dplyr::pull()

# read in data
bgc_pts_subzone <- readRDS(fs::path(out_dir , "model_input_pts.rds"))

run_balance_optimisation(bgc_pts_subzone,
                         fuzz_matrix,
                         covars,
                         extra_pts = TRUE,
                         out_dir,
                         ds_iterations = 10, # c(10, 20,  30, 40, 50, 60, 70, 80, 90),
                         smote_iterations = NA, #c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
                         use_neighbours = FALSE)





```


```{r}
#| eval: false

############################################################

## 4.3) Run the final model with the balance option

############################################################

out_dir = fs::path(PEMprepr::read_fid()$dir_3020_draft$path_rel, "30_nf")

fuzz_matrix <- read.csv(fs::path(out_dir , "fuzzy_matrix.csv" ))

covars = read.csv(fs::path(out_dir , "reduced_covariate_list.csv")) |> dplyr::pull()

bgc_pts_subzone <- readRDS(fs::path(out_dir , "model_input_pts.rds"))

model_bal = "aspat_paf_theta0"

# "aspat_paf_theta.5" ,"aspat_paf_theta0" ,
# "aspat_paf_theta1" , "aspatial_sum", "spat_paf_theta.5" , "spat_paf_theta0", "spat_paf_theta1",
#  "spatial_sum", "overall"

run_final_model(
  train_data = bgc_pts_subzone,
  covars = covars,
  model_bal = model_bal,
  report = FALSE,
  out_dir = out_dir,
  extra_pts = TRUE,
  ds_ratio = NA,
  sm_ratio = NA
)


######################
# 4) Predict maps -  balanced model
######################

model_dir = fs::path(PEMprepr::read_fid()$dir_3020_draft$path_rel, "30_nf")
covars = read.csv(fs::path(model_dir, "reduced_covariate_list.csv")) |>  dplyr::pull()
cov_dir = fs::path(PEMprepr::read_fid()$dir_1020_covariates$path_rel, "5m")
tile_dir = fs::path(PEMprepr::read_fid()$dir_30_model$path_rel,"tiles")
bec_shp = sf::st_read(fs::path(PEMprepr::read_fid()$dir_1010_vector$path_rel,"bec.gpkg"), quiet = TRUE)
model_name = "final_model_aspat_paf_theta0.rds"

run_predict_map(
  model_type = "nf",
  model_dir = model_dir,
  model_name = model_name,
  covars = covars,
  cov_dir = cov_dir,
  tile_dir = tile_dir,
  bec_shp = bec_shp
)


```





