---
title: "Setup-new-aoi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup-new-aoi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## 1. Setting up a new PEM project 

The PEMr package is designed to help users access the functions and follow a workflow to create a Predictive Ecosystem Map project. 

These packages are currently in development and there maybe breaking changes. 

The first step needed is to generate a new PEM project. This will include a new R-studio project, a template folder structure, and a series of template workflow quarto documents. 


```{r}
#| eval: false
remotes::install_github("ninoxconsulting/PEMr", build_vignettes = TRUE)
remotes::install_github("ninoxconsulting/PEMprepr", build_vignettes = TRUE)

library(PEMr)
library(PEMprepr)

```

The first step is to decide the name of your area of interest (AOI) and if you have a spatial file for the given area. This is not required at this stage but if a convenient feature. 


```{r}
#| eval: false
#| 
create_pemr_project(
  path = "C:\\r_repo\\2024_pemr\\testing_pemr",
  aoi = "datecreek_aoi",
  aoi_file = "C:\\r_repo\\2024_pemr\\datecreek_aoi.gpkg"
)


```



## 2. Generating base data for modelling (5m resolution as default) 

```{r setup, eval = FALSE}
library(PEMr)
library(PEMprepr)

#project_name <- "{{ aoi_name }}"
#project_name <- "datecreek_aoi"

fid <- read_fid()

```


### create model templates 

```{r, eval = FALSE}

aoi_dir = read_fid()$dir_0010_vector$path_abs
files <- list.files(aoi_dir, pattern = "([.]gpkg)|([.]shp)$")
aoi_file = fs::path(aoi_dir, files )


# vector
aoi <- snap_aoi(aoi_file, method = "expand")

aoi <- fs::path(PEMprepr::read_fid()$dir_1010_vector$path_abs, "aoi_snapped.gpkg")

# raster

# Prepare Raster Data(for 5m lidar template)

r5 <- create_template_raster(aoi, res = 5)

```


### generate base vector layers

```{r, eval = FALSE}

create_base_vectors(aoi, out_dir = PEMprepr::read_fid()$dir_1010_vector$path_abs)

```


### generate base DEM layer for modelling (5m)

```{r, eval = FALSE}

#1) option to use trim data 

dem = get_cded_dem(aoi = fs::path(PEMprepr::read_fid()$dir_1020_covariates$path_abs, "25m", "template.tif"),
                   res = 5,
                   overwrite = TRUE)

#2) option to use lidr 


```

### generate raster covariates for modelling (5m)


```{r, eval = FALSE}

# get model covariates 

find_saga_path()

saga_cmd()


create_covariates(
   dtm = fs::path(PEMprepr::read_fid()$dir_1020_covariates$path_rel,"5m","dem.tif"),
   saga_path = saga_cmd()
   out_dir = PEMprepr::read_fid()$dir_1020_covariates$path_rel,
   layers = "all"
)


```


### Create a bec raster to match the template sizes 

```{r, eval = FALSE}
# generate a 5m raster
bec_rast5 <- create_bgc_template(
  field = "MAP_LABEL",
  template =  r5,
  write_output = FALSE
)


```


congratulations you are now ready to start developing the sample plan



## 3. Generating landscape data for sample plan developemtn (25m resolution as default) 


## this is a placeholder for the workflow of samplr (still in developement)

```{r, eval = FALSE}

# vector (aoi, roads and start points)

in_dir = PEMprepr::read_fid()$dir_1010_vector$path_abs
aoi_file = fs::path(in_dir, "aoi_snapped.gpkg" )

out_dir <- PEMprepr::read_fid()$dir_201010_inputs$path_abs


aoils <- snap_aoi(aoi_file, method = "expand", buffer = 1000, write_output = FALSE)

sf::st_write(aoils, fs::path(out_dir, "aoi_snapped_ls.gpkg"), append = FALSE)



# Prepare Raster Data (for landscape)

r25 <- create_template_raster(aoils,
                              res = 25,
                              out_dir = PEMprepr::read_fid()$dir_201010_inputs$path_abs,
                              write_output = TRUE)


dem25 = get_cded_dem(aoi = r25,
                     res = 25,
                     out_dir = PEMprepr::read_fid()$dir_201010_inputs$path_abs,
                     write_output = TRUE)



# generate a 25m raster

bec_rast25 <- create_bgc_template(
  field = "MAP_LABEL",
  template=  r25,
  out_dir = PEMprepr::read_fid()$dir_201010_inputs$path_abs,
  write_output = TRUE
)



# Generate landscape covariates

create_landscape_covariates(
  dtm = dem25,
  saga_path = "C:/Programs/saga-9.2.0_x64/saga-9.2.0_x64/saga_cmd.exe",
  layers = c("mrvbf", "dah", "landform"),
  out_dir = PEMprepr::read_fid()$dir_201010_inputs$path_abs,
  sieve_size = 10,
  dah_threshold = 0.2,
  saga_param = list(
    T_SLOPE = 64, TPCTL_V = 6, T_PCTL_R = 2,
    P_SLOPE = 4.0, P_PCTL = 3.0, UPDATE = 1,
    CLASSIFY = 1, MAX_RES = 100
  ))



# create binned landscape
landscapes <- create_binned_landscape(
  in_dir = fs::path(PEMprepr::read_fid()$dir_201010_inputs$path_abs,"25m","modules_landscape"),
  layers = c("dah_LS", "landform_LS","mrvbf_LS"),
  write_output = TRUE)

terra::plot(landscapes)


# check landscape class with Bec
routdf <- check_bgc_landscapes(bec_rast25, landscapes)

ggplot2::ggplot(routdf, ggplot2::aes(landscape)) +
  ggplot2::geom_histogram() +
  ggplot2::facet_wrap(~MAP_LABEL)


# download roads layer and run through standard checks

out_dir <- PEMprepr::read_fid()$dir_201010_inputs$path_abs

roadsls <- get_roads(aoils, out_dir)

check_roads(roadsls)




# Get start points and run checks

cities <- sf::st_read(fs::path(PEMprepr::read_fid()$dir_1010_vector$path_rel,"major_towns_bc.gpkg"))

# not in study area
nearest_town = "Hazelton"
start1 <- cities[cities$NAME == nearest_town,"NAME"]

check_locations( roads = roadsls, locations = start1)

# in study area
nearest_town = "TEST_location"
start <- cities[cities$NAME == nearest_town,"NAME"]
check_locations( roads = roadsls, locations = start)

# one in and one out of study area
start2 <- rbind(start, start1)
check_locations( roads = roadsls, locations = start2)



sf::st_write(start2, fs::path(out_dir, "start_ls.gpkg"), append = FALSE)


## Generate the cost layer - need to review these two scripts and 


## Currently there are two options
## old version
#03_sample_plan_new_version.R

## new version
#03_sample_plan_new_version.R
#fn_prep_cost_layer_utils.R

```





