---
title: "Setup-new-aoi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup-new-aoi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## 1. Setting up a new PEM project 

The PEMr package is designed to help users access the functions and follow a workflow to create a Predictive Ecosystem Map project. 

These packages are currently in development and there maybe breaking changes. 

The first step needed is to generate a new PEM project. This will include a new R-studio project, a template folder structure, and a series of template workflow quarto documents. 


```{r}
#| eval: false
remotes::install_github("ninoxconsulting/PEMr", build_vignettes = TRUE)
remotes::install_github("ninoxconsulting/PEMprepr", build_vignettes = TRUE)

library(PEMr)
library(PEMprepr)

```

The first step is to decide the name of your area of interest (AOI) and if you have a spatial file for the given area. This is not required at this stage but if a convenient feature. 


```{r}
#| eval: false
#| 
create_pemr_project(
  path = "C:\\r_repo\\2024_pemr\\testing_pemr",
  aoi = "datecreek_aoi",
  aoi_file = "C:\\r_repo\\2024_pemr\\datecreek_aoi.gpkg"
)


```




